\MakeAbstract{
    本文面向双轮足机器人在平坦地面高效移动与复杂地形稳定通过之间的兼顾需求，围绕混合式腿部结构设计与轻量级控制方法开展研究。针对传统四足轮腿机器人存在执行器数量多、结构复杂、控制计算负担较重等问题，本文选择双轮足机器人作为研究对象，旨在在保证运动性能的同时实现较好的结构简化与嵌入式实时控制能力。

    在机械结构设计方面，围绕轻量化、高动态、高刚度和高集成的总体目标，完成了整机机械方案设计，明确了自由度配置、性能指标及关键尺寸参数；进一步完成了跳跃系统、腿部传动系统和移动系统设计，并对关键执行器进行了选型与校核，为后续控制系统设计与样机实现提供了硬件基础。

    在建模与控制方法方面，针对混合式双轮足机器人腿部闭环机构耦合强、任务空间控制不直观的问题，本文基于虚功等效原理将单腿抽象为矢状面内的可伸缩虚拟杆，建立了整机平面等效运动学与动力学模型，并给出了五连杆机构在 VMC 框架下从任务空间到关节空间的力矩映射关系。在此基础上，提出了一种面向嵌入式实时实现的分层轻量级混合控制系统：矢状面主通道采用 LQR 实现平衡与速度跟踪，腿部任务空间采用 VMC 实现支撑与摆动调节，侧向稳定部分采用基于翻滚角几何关系的腿长差补偿方法，并在执行层引入轻量级 QP 进行约束投影，以兼顾控制性能、结构清晰性与在线计算效率。

    在验证方面，围绕原地平衡、速度跟踪、抗扰恢复、上下坡与越障以及转向稳定等典型工况，分别开展了仿真分析与样机实机验证。结果表明，本文所提出的混合式结构与分层轻量级控制系统能够有效提升双轮足机器人在多种工况下的稳定运动能力，验证了所提方法的有效性与工程可实现性。
}{双轮足机器人，混合式结构，运动学与动力学建模，分层控制，轻量级控制}

\MakeAbstractEng{
    This thesis investigates the structural design and lightweight control of a hybrid wheeled-biped robot for achieving both efficient locomotion on flat ground and stable traversal in complex terrains. To address the problems of high actuator redundancy, structural complexity, and heavy computational burden in conventional wheeled-legged robots, a two-wheeled biped robot is selected as the research platform, with the goal of balancing locomotion capability, structural simplicity, and embedded real-time control feasibility.

    In terms of mechanical design, oriented toward lightweight design, high dynamic performance, high stiffness, and high integration, the overall mechanical scheme of the robot is developed, including degree-of-freedom allocation, target performance indices, and key dimensional parameters. The jumping subsystem, leg transmission subsystem, and mobile subsystem are further designed, and key actuators are selected and verified, providing the hardware foundation for controller implementation and prototype development.

    In terms of modeling and control, to address the strong coupling of the hybrid leg closed-chain mechanism and the lack of intuitive task-space description, each leg is equivalently modeled as a telescopic virtual leg in the sagittal plane based on the virtual-work principle. On this basis, the planar equivalent kinematic and dynamic models of the whole robot are established, and the torque mapping of the five-bar leg mechanism from task space to joint space under the VMC framework is derived. A layered lightweight hybrid control system suitable for embedded real-time implementation is then proposed. In the sagittal plane, an LQR controller is adopted for balance and velocity tracking; in the leg task space, VMC is used for support and swing regulation; for lateral stabilization, a geometric roll-angle-based leg-length-difference compensation method is introduced; and a lightweight QP is employed in the execution layer for constraint projection. This design jointly considers control performance, structural clarity, and online computational efficiency.

    Finally, simulation studies and prototype experiments are conducted under typical scenarios including standing balance, velocity tracking, disturbance recovery, slope and obstacle negotiation, and turning stabilization. The results demonstrate that the proposed hybrid structure and layered lightweight control system can effectively improve the stability and motion capability of the wheeled-biped robot under multiple operating conditions, thereby validating the effectiveness and engineering feasibility of the proposed method.
}{wheeled-biped robot, hybrid mechanism, kinematic and dynamic modeling, layered control, lightweight control}
